---
title: "Prediction Assignment Writeup"
output: html_document
---

```{r setup, include=FALSE,warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Overview
The aim of this project is to predict the manner in which 6 participants performed some exercises. The machine learning algorithm described here is applied to the 20 test cases available in the test data and the predictions are submitted in appropriate format to the Course Project Prediction Quiz for automated grading.

##Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: [http://groupware.les.inf.puc-rio.br/har]http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

####Data Loading and Exploratory Analysis

* Source of Data:

The training data for this project are available here:

[Training Set]https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

[Test Set]https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

* Loading Libraries:
```{r, warning=FALSE, message=FALSE}
library(knitr)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
library(caret)
set.seed(1234)
```

##### Loading and cleaning the data:

```{r}
# Set the URLs
TrainDataURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
TestDataURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# Download the datasets
training <- read.csv(TrainDataURL)
testing <- read.csv(TestDataURL)
```

```{r}
dim(training)
```

```{r}
dim(testing)
```

```{r}
# create a partition using caret with the training dataset on 70,30 ratio
inTrain  <- createDataPartition(training$classe, p=0.7, list=FALSE)

TrainSet <- training[inTrain, ]

TestSet  <- training[-inTrain, ]
dim(TrainSet)
```

```{r}
dim(TestSet)
```
Both created datasets have 160 variables. Let’s clean NA, The Near Zero variance (NZV) variables and the ID variables as well.
```{r}
# remove variables with Nearly Zero Variance
NZV <- nearZeroVar(TrainSet)
TrainSet <- TrainSet[, -NZV]
TestSet  <- TestSet[, -NZV]
dim(TestSet)
```

```{r}
dim(TrainSet)
```

```{r}
# remove variables that are mostly NA
AllNA    <- sapply(TrainSet, function(x) mean(is.na(x))) > 0.95
TrainSet <- TrainSet[, AllNA==FALSE]
TestSet  <- TestSet[, AllNA==FALSE]
dim(TestSet)
```

```{r}
dim(TrainSet)
```

```{r}
# removing identification only variables (columns 1 to 5)
TrainSet <- TrainSet[, -(1:5)]
TestSet  <- TestSet[, -(1:5)]
dim(TrainSet)
```

After cleaning, we can see that the number of vairables for the analysis are now only 53.

#####Coorelation Analysis
Analysing correlation among variables before proceeding to modeling procedures.

```{r}
corMatrix <- cor(TrainSet[, -54])
corrplot(corMatrix, order = "FPC", method = "color", type = "lower", 
         tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```

The highly correlated variables are shown in dark colors in the graph above. To make an even more compact analysis, a PCA (Principal Components Analysis) could be performed as pre-processing step to the datasets. Nevertheless, as the correlations are quite few, this step will not be applied for this assignment.

#####Prediction Model Building

Three popular methods will be applied to model the regressions (in the Train dataset) and the best one (with higher accuracy when applied to the Test dataset) will be used for the quiz predictions. The methods are: Random Forests, Decision Tree and Generalized Boosted Model, as described below. A Confusion Matrix is plotted at the end of each analysis to better visualize the accuracy of the models.

*Random Forests
```{r}
set.seed(1234)
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modFitRandForest <- train(classe ~ ., data=TrainSet, method="rf",
                          trControl=controlRF)
modFitRandForest$finalModel
```

```{r}
# prediction on Test dataset
predictRandForest <- predict(modFitRandForest, newdata=TestSet)
confMatRandForest <- confusionMatrix(predictRandForest, factor( TestSet$classe))
confMatRandForest
```

```{r}
# plot matrix results
plot(confMatRandForest$table,  
     main = paste("Random Forest - Accuracy =",
                  round(confMatRandForest$overall['Accuracy'], 4)))
```

*Decision Tree
```{r}
set.seed(301)
modFitDecTree <- rpart(classe ~ ., data=TrainSet, method="class")
fancyRpartPlot(modFitDecTree)
```

```{r}
# prediction on Test dataset
predictDecTree <- predict(modFitDecTree, newdata=TestSet, type="class")
confMatDecTree <- confusionMatrix(predictDecTree, factor(TestSet$classe))
confMatDecTree
```

```{r}
# plot matrix results
plot(confMatDecTree$table, col = confMatDecTree$byClass, 
     main = paste("Decision Tree - Accuracy =",
                  round(confMatDecTree$overall['Accuracy'], 4)))
```


*Generalized Boosted Model (GBM)
```{r}
# model fit
set.seed(301)
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
modFitGBM  <- train(classe ~ ., data=TrainSet, method = "gbm",
                    trControl = controlGBM, verbose = FALSE)
modFitGBM$finalModel
```

```{r}
# prediction on Test dataset
predictGBM <- predict(modFitGBM, newdata=TestSet)
confMatGBM <- confusionMatrix(predictGBM, factor(TestSet$classe))
confMatGBM
```

```{r}
# plot matrix results
plot(confMatGBM$table, col = confMatGBM$byClass, 
     main = paste("GBM - Accuracy =", round(confMatGBM$overall['Accuracy'], 4)))
```

##### Applying selected model to Test Data
Following is the accuracy of each regression modeling method:
*Random Forest : 0.9986 
*Decision Tree : 0.7295 
*GBM : 0.9884 
I
The Random Forest model will be applied to predict the 20 quiz results (testing dataset):
```{r}
predictTEST <- predict(modFitRandForest, newdata=testing)
predictTEST
```



